\cleardoublepage
\begin{thebibliography}{99}
\addcontentsline{toc}{chapter}{Bibliografía}

\bibitem{zeppenfeld2022}
Zeppenfeld, K. y colaboradores.
\newblock 2022 ESC Guidelines for the management of patients with ventricular arrhythmias and the prevention of sudden cardiac death.
\newblock \emph{European Heart Journal}, 43(40), 3997--4126, 2022.
\newblock \url{https://doi.org/10.1093/eurheartj/ehac262}.

\bibitem{kligfield2007}
Kligfield, P., Gettes, L. S., Bailey, J. J. y colaboradores.
\newblock Recommendations for the standardization and interpretation of the electrocardiogram. Part I. The electrocardiogram and its technology.
\newblock \emph{Circulation}, 115(10), 1306--1324, 2007.
\newblock \url{https://doi.org/10.1161/CIRCULATIONAHA.106.180200}.

\bibitem{surawicz2009}
Surawicz, B., Childers, R., Deal, B. J., Gettes, L. S. y colaboradores.
\newblock AHA, ACCF and HRS recommendations for the standardization and interpretation of the electrocardiogram. Part III. Intraventricular conduction disturbances.
\newblock \emph{Circulation}, 119(10), e235--e240, 2009.
\newblock \url{https://doi.org/10.1161/CIRCULATIONAHA.108.191095}.

\bibitem{rautaharju2009st}
Rautaharju, P. M., Surawicz, B. y Gettes, L. S.
\newblock AHA, ACCF and HRS recommendations for the standardization and interpretation of the electrocardiogram. Part IV. The ST segment, T and U waves, and the QT interval.
\newblock \emph{Circulation}, 119(10), e241--e250, 2009.
\newblock \url{https://doi.org/10.1161/CIRCULATIONAHA.108.191096}.

\bibitem{sieira2017}
Sieira, J. y Brugada, P.
\newblock The definition of the Brugada syndrome.
\newblock \emph{European Heart Journal}, 38(40), 3029--3034, 2017.
\newblock \url{https://doi.org/10.1093/eurheartj/ehx490}.

\bibitem{wilde2023scbt}
Wilde, A. A. M., Semsarian, C., Marquez, M. F. y colaboradores.
\newblock Use, misuse, and pitfalls of the drug challenge test in the diagnosis of the Brugada syndrome.
\newblock \emph{European Heart Journal}, 44(27), 2427--2439, 2023.
\newblock \url{https://doi.org/10.1093/eurheartj/ehad295}.

\bibitem{brugadadrugsdiagnostic}
BrugadaDrugs.org Advisory Board.
\newblock Diagnostic drugs for Brugada syndrome.
\newblock BrugadaDrugs.org.
\newblock \url{https://www.brugadadrugs.org/druglist_diagnostic/}.

\bibitem{ecgpediaType1Image}
CardioNetworks y ECGpedia.
\newblock Brugada syndrome type1 example4.
\newblock Wikimedia Commons, licencia CC BY-SA 3.0.
\newblock \url{https://commons.wikimedia.org/wiki/File:Brugada_syndrome_type1_example4_(CardioNetworks_ECGpedia).png}.

\bibitem{napolitano2006}
Napolitano, C. y Priori, S. G.
\newblock Brugada syndrome.
\newblock \emph{Orphanet Journal of Rare Diseases}, 1, 35, 2006.
\newblock \url{https://doi.org/10.1186/1750-1172-1-35}.

\bibitem{corrado2010}
Corrado, D., Pelliccia, A., Heidbuchel, H. y colaboradores.
\newblock Recommendations for interpretation of 12 lead electrocardiogram in the athlete.
\newblock \emph{European Heart Journal}, 31(2), 243--259, 2010.
\newblock \url{https://doi.org/10.1093/eurheartj/ehp473}.

\bibitem{ribeiro2020}
Ribeiro, A. H. y colaboradores.
\newblock Automatic diagnosis of the 12 lead ECG using a deep neural network.
\newblock \emph{Nature Communications}, 11, 1760, 2020.
\newblock \url{https://doi.org/10.1038/s41467-020-15432-4}.

\bibitem{wagner2020}
Wagner, P., Strodthoff, N., Bousseljot, R. D. y colaboradores.
\newblock PTB XL, a large publicly available electrocardiography dataset.
\newblock \emph{Scientific Data}, 7, 154, 2020.
\newblock \url{https://doi.org/10.1038/s41597-020-0495-6}.

\bibitem{gow2023}
Gow, B., Pollard, T., Nathanson, L. A. y colaboradores.
\newblock MIMIC IV ECG. Diagnostic Electrocardiogram Matched Subset. Version 1.0.
\newblock PhysioNet, 2023.
\newblock \url{https://doi.org/10.13026/4nqg-sb35}.

\bibitem{gliner2020}
Gliner, V., Keidar, N., Makarov, V. y colaboradores.
\newblock Automatic classification of healthy and disease conditions from images or digital standard 12 lead electrocardiograms.
\newblock \emph{Scientific Reports}, 10, 16331, 2020.
\newblock \url{https://doi.org/10.1038/s41598-020-73060-w}.

\bibitem{sangha2022}
Sangha, V. y colaboradores.
\newblock Automated multilabel diagnosis on electrocardiographic images and signals.
\newblock \emph{Nature Communications}, 13, 1583, 2022.
\newblock \url{https://doi.org/10.1038/s41467-022-29153-3}.

\bibitem{wu2022digitization}
Wu, H., Patel, K. H. K., Li, X. y colaboradores.
\newblock A fully automated paper ECG digitisation algorithm using deep learning.
\newblock \emph{Scientific Reports}, 12, 20963, 2022.
\newblock \url{https://doi.org/10.1038/s41598-022-25284-1}.

\bibitem{gliner2023}
Gliner, V., Makarov, V., Avetisyan, A. I., Schuster, A. y Yaniv, Y.
\newblock Using domain adaptation for classification of healthy and disease conditions from mobile captured images of standard 12 lead electrocardiograms.
\newblock \emph{Scientific Reports}, 13, 14023, 2023.
\newblock \url{https://doi.org/10.1038/s41598-023-40693-6}.

\bibitem{sayadi2010}
Sayadi, O., Shamsollahi, M. B. y Clifford, G. D.
\newblock Synthetic ECG generation and Bayesian filtering using a Gaussian wave based dynamical model.
\newblock \emph{Physiological Measurement}, 31(10), 1309--1329, 2010.
\newblock \url{https://doi.org/10.1088/0967-3334/31/10/002}.

\bibitem{gillette2023}
Gillette, K. y colaboradores.
\newblock MedalCare XL. 16 900 healthy and pathological synthetic 12 lead ECGs from electrophysiological simulations.
\newblock \emph{Scientific Data}, 10, 531, 2023.
\newblock \url{https://doi.org/10.1038/s41597-023-02416-4}.

\bibitem{alcaraz2023}
Lopez Alcaraz, J. M. y Strodthoff, N.
\newblock Diffusion based conditional ECG generation with structured state space models.
\newblock arXiv:2301.08227, 2023.
\newblock \url{https://doi.org/10.48550/arXiv.2301.08227}.

\bibitem{liu2026}
Liu, R., Bai, Y., Yue, X. y Zhang, P.
\newblock Teaching multimodal language models to comprehend 12 lead electrocardiographic images.
\newblock \emph{npj Digital Medicine}, 9, 349, 2026.
\newblock \url{https://doi.org/10.1038/s41746-026-02551-3}.

\bibitem{zhang2026meeti}
Zhang, D., Lan, X., Geng, S. y colaboradores.
\newblock MEETI. A multimodal ECG dataset from MIMIC IV ECG with signals, images, features and interpretations.
\newblock \emph{Scientific Data}, 13, 527, 2026.
\newblock \url{https://doi.org/10.1038/s41597-026-06796-1}.

\bibitem{alayrac2022}
Alayrac, J. B., Donahue, J., Luc, P. y colaboradores.
\newblock Flamingo. A visual language model for few shot learning.
\newblock \emph{Advances in Neural Information Processing Systems}, 35, 23716--23736, 2022.
\newblock \url{https://arxiv.org/abs/2204.14198}.

\bibitem{chen2024}
Chen, S., Han, Z., He, B., Liu, J., Buckley, M., Qin, Y., Torr, P., Tresp, V. y Gu, J.
\newblock Can multimodal large language models truly perform multimodal in context learning.
\newblock \emph{IEEE/CVF Winter Conference on Applications of Computer Vision}, 6000--6010, 2025.
\newblock \url{https://doi.org/10.1109/WACV61041.2025.00585}.

\bibitem{zhang2024mmicl}
Qin, L., Chen, Q., Fei, H., Chen, Z., Li, M. y Che, W.
\newblock What factors affect multimodal in context learning. An in depth exploration.
\newblock \emph{Advances in Neural Information Processing Systems}, 37, 2024.
\newblock \url{https://arxiv.org/abs/2410.20482}.

\bibitem{camba2026icl}
Camba, S., Otero, A., García, D., Sánchez, L. y Costa, N.
\newblock Can In-Context Learning Enable Large Vision Language Models to Detect ECG Abnormalities?
\newblock En \emph{Artificial Intelligence in Biomedicine. CIABiomed 2025}, Lecture Notes in Computer Science, vol. 16148, 200--213. Springer, Cham, 2026.
\newblock \url{https://doi.org/10.1007/978-3-032-10661-2_16}.

\bibitem{gemma4card}
Google DeepMind.
\newblock Gemma 4 model card.
\newblock Google AI for Developers.
\newblock \url{https://ai.google.dev/gemma/docs/core/model_card_4}.

\bibitem{medgemma15}
Google Health AI Developer Foundations.
\newblock MedGemma 1.5 model card.
\newblock Google for Developers.
\newblock \url{https://developers.google.com/health-ai-developer-foundations/medgemma/model-card}.

\bibitem{he2016}
He, K., Zhang, X., Ren, S. y Sun, J.
\newblock Deep residual learning for image recognition.
\newblock \emph{Proceedings of the IEEE Conference on Computer Vision and Pattern Recognition}, 770--778, 2016.
\newblock \url{https://doi.org/10.1109/CVPR.2016.90}.

\bibitem{selvaraju2017}
Selvaraju, R. R. y colaboradores.
\newblock Grad CAM. Visual explanations from deep networks via gradient based localization.
\newblock \emph{Proceedings of the IEEE International Conference on Computer Vision}, 618--626, 2017.
\newblock \url{https://doi.org/10.1109/ICCV.2017.74}.

\bibitem{ganin2016}
Ganin, Y. y colaboradores.
\newblock Domain adversarial training of neural networks.
\newblock \emph{Journal of Machine Learning Research}, 17(59), 1--35, 2016.
\newblock \url{https://jmlr.org/papers/v17/15-239.html}.

\bibitem{tsoumakas2007}
Tsoumakas, G. y Katakis, I.
\newblock Multi label classification. An overview.
\newblock \emph{International Journal of Data Warehousing and Mining}, 3(3), 1--13, 2007.
\newblock \url{https://doi.org/10.4018/jdwm.2007070101}.

\bibitem{pytorchrepro}
PyTorch Contributors.
\newblock Reproducibility.
\newblock PyTorch documentation.
\newblock \url{https://docs.pytorch.org/docs/stable/notes/randomness.html}.

\bibitem{pytorchbce}
PyTorch Contributors.
\newblock BCEWithLogitsLoss.
\newblock PyTorch documentation.
\newblock \url{https://docs.pytorch.org/docs/stable/generated/torch.nn.BCEWithLogitsLoss.html}.

\bibitem{torchvisionresnet18}
Torchvision Contributors.
\newblock ResNet18 and pretrained weights.
\newblock Torchvision documentation.
\newblock \url{https://docs.pytorch.org/vision/stable/models/generated/torchvision.models.resnet18.html}.

\bibitem{loshchilov2019}
Loshchilov, I. y Hutter, F.
\newblock Decoupled weight decay regularization.
\newblock \emph{International Conference on Learning Representations}, 2019.
\newblock \url{https://arxiv.org/abs/1711.05101}.

\bibitem{vllmserver}
vLLM Project.
\newblock OpenAI compatible server.
\newblock vLLM documentation.
\newblock \url{https://docs.vllm.ai/en/latest/serving/online_serving/openai_compatible_server/}.

\bibitem{vllmgemma4}
vLLM Project.
\newblock Gemma 4 usage guide.
\newblock vLLM recipes.
\newblock \url{https://docs.vllm.ai/projects/recipes/en/latest/Google/Gemma4.html}.

\bibitem{costacortez2026brugada}
Costa Cortez, N. y García Iglesias, D.
\newblock Brugada-HUCA. 12-Lead ECG Recordings for the Study of Brugada Syndrome. Version 1.0.0.
\newblock PhysioNet, 2026.
\newblock \url{https://doi.org/10.13026/0m2w-dy83}.

\bibitem{collins2024tripodai}
Collins, G. S., Moons, K. G. M., Dhiman, P. y colaboradores.
\newblock TRIPOD+AI statement. Updated guidance for reporting clinical prediction models that use regression or machine learning methods.
\newblock \emph{BMJ}, 385, e078378, 2024.
\newblock \url{https://doi.org/10.1136/bmj-2023-078378}.

\bibitem{moons2025probastai}
Moons, K. G. M., Damen, J. A. A. G., Kaul, T. y colaboradores.
\newblock PROBAST+AI. An updated quality, risk of bias and applicability assessment tool for prediction models using regression or artificial intelligence methods.
\newblock \emph{BMJ}, 388, e082505, 2025.
\newblock \url{https://doi.org/10.1136/bmj-2024-082505}.

\bibitem{gdpr}
Unión Europea.
\newblock Reglamento 2016/679 relativo a la protección de las personas físicas en lo que respecta al tratamiento de datos personales.
\newblock \emph{Diario Oficial de la Unión Europea}.
\newblock \url{https://eur-lex.europa.eu/eli/reg/2016/679/oj/spa}.

\bibitem{lopdgdd}
España.
\newblock Ley Orgánica 3/2018 de Protección de Datos Personales y garantía de los derechos digitales.
\newblock \emph{Boletín Oficial del Estado}.
\newblock \url{https://www.boe.es/buscar/act.php?id=BOE-A-2018-16673}.

\bibitem{ley142007}
España.
\newblock Ley 14/2007 de Investigación biomédica.
\newblock \emph{Boletín Oficial del Estado}.
\newblock \url{https://www.boe.es/buscar/act.php?id=BOE-A-2007-12945}.

\bibitem{aepd2021risk}
Agencia Española de Protección de Datos.
\newblock Gestión del riesgo y evaluación de impacto en tratamientos de datos personales.
\newblock AEPD.
\newblock \href{https://www.aepd.es/guias/gestion-riesgo-y-evaluacion-impacto-en-tratamientos-datos-personales.pdf}{Documento oficial de la AEPD}.

\bibitem{aiact2024}
Unión Europea.
\newblock Reglamento 2024/1689 por el que se establecen normas armonizadas en materia de inteligencia artificial.
\newblock \emph{Diario Oficial de la Unión Europea}.
\newblock \url{https://eur-lex.europa.eu/eli/reg/2024/1689/oj/spa}.

\bibitem{mdr2017}
Unión Europea.
\newblock Reglamento 2017/745 sobre los productos sanitarios.
\newblock \emph{Diario Oficial de la Unión Europea}.
\newblock \url{https://eur-lex.europa.eu/eli/reg/2017/745/oj/spa}.

\bibitem{mdcg2019rev1}
Medical Device Coordination Group.
\newblock MDCG 2019-11 Rev. 1. Qualification and classification of software. Regulation (EU) 2017/745 and Regulation (EU) 2017/746.
\newblock Comisión Europea.
\newblock \url{https://health.ec.europa.eu/latest-updates/update-mdcg-2019-11-rev1-qualification-and-classification-software-regulation-eu-2017745-and-2025-06-17_en}.

\bibitem{mdcg2025six}
Medical Device Coordination Group.
\newblock MDCG 2025-6. FAQ on interplay between the Medical Devices Regulation, the In Vitro Diagnostic Medical Devices Regulation and the Artificial Intelligence Act.
\newblock Comisión Europea.
\newblock \url{https://health.ec.europa.eu/latest-updates/mdcg-2025-6-faq-interplay-between-medical-devices-regulation-vitro-diagnostic-medical-devices-2025-06-19_en}.

\bibitem{ehds2025}
Unión Europea.
\newblock Reglamento 2025/327 relativo al Espacio Europeo de Datos de Salud.
\newblock \emph{Diario Oficial de la Unión Europea}.
\newblock \url{https://eur-lex.europa.eu/eli/reg/2025/327/oj/spa}.

\end{thebibliography}
