\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
\markboth{Abstract}{Abstract}

This Bachelor's Thesis addresses automatic ECG analysis from ECG images, understood here as visual representations of the trace rather than the original temporal signals. The case study is suspected Brugada syndrome, a rare disease whose low incidence makes it especially difficult to model with few data. The goal is to study how a small number of labelled examples can support the review of ECGs with compatible morphology.

The thesis compares two ways of dealing with this scarcity of labelled data: a conventional convolutional approach and a vision-language approach based on in-context learning. Both approaches receive ECGs as images, are evaluated at patient level and use the same example-set sizes \(K=\{2,4,8,16,32\}\). The evaluation combines two data sources. The first is a QRS/ST morphological simulator that generates one V1 image per simulated patient and assigns labels using rules on trace morphology. The second is a real dataset from the Hospital Universitario Central de Asturias (HUCA), with 317 patients and one V1 image per patient associated with a patient-level clinical label.

The first approach is supervised. A ResNet18 uses labelled simulator patients to update its weights and learn three morphological findings related to the pathology under study. A supervised real-to-real CNN is not trained with the real HUCA dataset because the clinical cohort does not contain independent image-level annotations for RBBB, ST elevation and T-wave inversion. The real evaluation therefore measures transfer from the simulator to the clinical cohort.

The second approach is multimodal. Gemma~4 is used as a vision-language model with in-context learning: the labelled examples do not modify the model weights, but are inserted into the prompt as demonstrations before the image to be classified. Zero-shot, normal ICL and balanced ICL are compared.

The results show different behaviour in each domain. On the simulator, the CNN reaches balanced accuracy 0.848 with 32 training examples, sensitivity 0.883 and specificity 0.812. On the real dataset, baseline performance remains close to chance, with a best balanced accuracy of 0.516 at \(K=16\). Domain adaptation improves this modestly to 0.549 with MMD at \(K=32\). In the VLM approach, demonstrations clearly improve over zero-shot on the simulator and reach balanced accuracy 0.754 with balanced context at \(K=8\). On real ECGs, the best point is around 0.522 at \(K=16\). Balanced context increases sensitivity to 0.713, but specificity falls to 0.332.

Overall, the work shows that ECG-as-image analysis contains useful signal in a synthetic domain, but transfer to real Brugada data remains limited with few examples. The CNN benefits more clearly from additional data on the simulator. In-context learning does not require retraining the model for each new dataset and is easier to move between evaluation settings, although in these experiments it increases sensitivity at the cost of lower specificity. When the VLM receives simulator context and is queried on the real HUCA ECGs, performance collapses back towards the negative class, reinforcing the difficulty of the synthetic-to-real gap. Neither approach is yet robust enough for reliable clinical triage. The main contribution of the thesis is to characterize this limitation through a reproducible protocol and to show that further progress will require more positive cases, clinical morphological annotations and external validation.

\textbf{Keywords:} ECG as image, Brugada syndrome, few labelled data, supervised learning, convolutional neural networks, vision-language models, in-context learning, clinical triage.
